Dependent Object Types

### 2012-08-05 article.tex
original dot-sub

    \end{multicols}
}

>>>>

    \end{multicols}
} % END dot-sub

\subsection*{Subtyping}

Figure~\ref{dot-sub} defines the subtyping judgement $\Gamma \ts S
\sub T$ which states that in environment $\Gamma$ type $S$ is a
subtype of type $T$.  Any term of type $S$ can then be regarded via
the subsumption rule (\textsc{sub}) as a term of type $T$.

As usual, subtyping is reflexive (\textsc{refl}) but a corresponding
rule for transitivity is missing.  However, most of the rules for
specific types in Figure~\ref{dot-sub} have transitivity built in. The
issue of transitivity is further discussed below.

To prove that a type $S$ is a subtype of a refinement type $T \refine
z {\seq D}$, it must hold that $S$ is a subtype of $T$ and that $S$
contains declarations that subsume the declarations $\seq D$
(\textsc{$\sub$-rfn}). The refinement type itself is a subtype of its
parent type $T$ and all its supertypes (\textsc{rfn-$\sub$}).

A type selection $p.L$ is a subtype of (all supertypes of) the upper bound $U$ of
$L$ in $p$ (\textsc{tsel-$\sub$}).
It is a supertype of (all subtypes of) its lower
bound $S$ (\textsc{$\sub$-tsel}).  Function types are subject to the
standard co-/contravariant subtyping rule 
\ifdep
 (\textsc{$\sub$-$\Rightarrow$}).
\else
  (\textsc{$\sub$-$\tfun$}).
\fi
%A singleton type $p\sing$ is a subtype of any underlying type of path
%$p$ (\textsc{single-$\sub$}). 
The final six rules in
Figure~\ref{dot-sub} turn the subtyping relation into a lattice with 
meets $\tand$, joins $\tor$, bottom element $\Bot$ and top element $\Top$.

One interesting aspect of the subtyping relation of DOT is that it is
reflexive (\textsc{refl}) but not transitive. Instead of a separate
rule that specifies global transitivity, we find transitivity encoded
in most of the inference rules of Figure~\ref{dot-sub}. There is only
a single situation where transitivity does not hold:  Given a type
declaration $\Ldecl L S U$ which is a member of some type $T$, and a
variable $x$ of type $T$, we have that $S \sub x.L$ by
(\textsc{$\sub$-tsel}) and $x.L \sub U$ by (\textsc{tsel-$\sub$}).
However, this does not imply that $S \sub U$. In fact, 
$T$ might be an {\em unrealizable} type, which admits no solution for
its $L$ member, precisely because the lower bound of $L$ is not a
subtype of its upper bound.

The type assignment rule (\textsc{new}) in Figure~\ref{dot-one}
has as one of its preconditions that the type of
the created object must be realizable: each lower bounds must be a subtype
of its corresponding upper bound. Subtyping can be shown to be transitive in DOT if the
least type of a subtype chain is realizable.
